rbcL ASV clustering

### 2019-11-02 code/rbcl_asv_clustering.Rmd
removed Chlainomonas high OTU

```{r read, include=F}
asv_key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
```

```{r asv_distance_matrix, include=F}

aligns <- asv_key %>%
  filter(asv_id %in% rel_abund$asv_id) %>% # only want asvs used in the main analysis
  deframe() %>% 
  DNAStringSet() %>% 
  DECIPHER::AlignSeqs(verbose=F)

# check alignments
aligns %>% DECIPHER::BrowseSeqs(highlight = T) # ASV 639 has a ~70 bp insert, 646 ~45 bp deletion, 535 ~110 bp deletion
length(aligns)

@@

ggplot(tsne_pts_taxonomy, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, is.na(plot_taxa)), 
             color = "grey", alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
             alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few()
```

@@

Manually edit the clustering:
  - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant
  - split *Chlainomonas* into two, based on the observation in IQTree the two clades of *Chlainomonas* seperate out by elevation

```{r edit-otus, echo=F, fig.width=8, fig.height=6}
tsne_edited_otus <- tsne_taxa_otu %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

tsne_edited_otus %>% 
  select(plot_taxa, otu) %>% 
  distinct() %>% 
  arrange(otu)

# name each OTU
otu_names <- tibble(otu = as.character(1:10), 
                    otu_name = c("Chloromonas spp. C",
                                 "Sanguina A","Chlainomonas B",
                                 "Chloromonas muramotoi E",
                                 "Chloromonas G",
                                 "Chloromonas I",
                                 "Algae H",
                                 "Trebouxiophyceae J",
                                 "Chlainomonas F",
                                 "Chloromonas krienitzii D"))

tsne_otu_names <- tsne_edited_otus %>% 
  left_join(otu_names, by="otu")

@@

ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_otu_names, is.na(plot_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_otu_names, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few() +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), show.legend = FALSE)
```

```{r include_rel_abund, echo=F, fig.width=8, fig.height=6}
total_rel_abund <- rel_abund %>% 
  group_by(asv_id) %>% 
  summarise(global_rel_abund = sum(rel_abund))

@@

final_tsne <- ggplot(tsne_rel_abund, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), 
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank", 
                             expand = unit(2, "mm")) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
             aes(size = global_rel_abund),
             color = "grey",alpha = 0.7) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order),
                 size = global_rel_abund), 
             alpha = 0.7) +
  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
  ggthemes::theme_few() +
  scale_size(range = c(1, 6), guide="none") ################################################# not changing the scale size
final_tsne
```

```{r tsne-output, echo=F}

# finel_tsne +
#   theme(legend.position = "none")
# 
# # extract the legend
# tsne_plot %>% 
#   get_legend() %>% 
#   as_ggplot 
# 
# ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
# ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
```

```{r asv_mds, echo=F, fig.width=10, fig.height=8}
mds_coords <- dists %>% 
  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
  as_tibble() %>% 
  add_column(asv_id = colnames(dists), .before=1)

@@

ggplot(mds_asv_plot, aes(V1, V2)) + 
  ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
                               filter(otu_name != "Trebouxiophyceae J"), # dont plot the ellipse around other treb
                             aes(fill = otu_name, label=otu_name),
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank",
                             expand = unit(2, "mm")) +
  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
  labs(title = "MDS clustering rbcL ASVs", color="ASV taxonomic assignment") +
  scale_size(range = c(1, 6), guide="none") +
  scale_alpha(guide="none")+
  theme_few() #+
  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
```

```{r write_mds, include=F}

@@

Write fasta for IQTree web serv
```{r}
asv.aligns %>% writeXStringSet(here("output","tree","rbcl","rbcl_aligns.fasta"))
```

Run IQtree
```{r tree, fig.width=8, fig.height=20}
rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))

idtax %>% filter(species == "Raphidonema_nivale")
# re-root tree
rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)

@@

# Plot trees
rbcl_ggtree %<+% select(tsne, asv_id, display_taxa) + # attach taxonomic assignment for tip label
  geom_tippoint(aes(color=display_taxa, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

rbcl_ggtree %<+% asv_otu_key + # attach OTU
  geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

# color edge red if boot < 80
# ggtree_conf <- rbcl_ggtree$data %>% 
#   separate(label, c("boot", "bayes"), sep = "/") %>% #split label into boots and bayes vals
#   mutate(insignificant_edge = case_when((boot < 80 & isTip == FALSE)~"dotted", 
#                                         TRUE~"solid"),
#          asv_id = case_when(isTip == TRUE~boot,
#                             TRUE~NA_character_)) %>% # round boot to nearest integer
#   left_join(otu_tbl, by="asv_id")

# rbcl_ggtree %<+% ggtree_conf +
#   aes(linetype = I(insignificant_edge)) +
#   geom_tippoint(aes(color=otu),alpha=0.7, size=3) + #
#   theme(legend.position = c(0.2,0.7), 
#         legend.title = element_blank(), 
#         legend.key = element_blank()) 
# ggsave("figs/ggtree.pdf", height = 10, width=7.5)
```

```{r output}

write_csv(asv_otu_key, here("output","keys","asv_otu_key.csv"))
```

>>>>

```{r read, include=F}

rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))

aligns <- readDNAStringSet(here("output/alignments/rbcl_aligns_in_main.fasta"))
tree <- read.tree(file=here("output/tree/rbcl/asv613.fasta.treefile")) # this is the tree with only the ASVs that appear in the filtered rel abund dataset (not including protocol stuff, etc)
```

```{r asv_distance_matrix, include=F}
BrowseSeqs(aligns, highlight = T)

@@

ggplot(tsne_pts_taxonomy, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, is.na(plot_taxa)), 
             color = "grey", alpha = 0.8) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
             alpha = 0.8) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment")
```

@@

Manually edit the clustering:
  - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant

```{r edit-otus, echo=F}
tsne_edited_otus <- tsne_taxa_otu %>%
  mutate(otu = case_when((V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "9",# krienitzii
                         TRUE ~ otu))

otus <- tsne_edited_otus %>% 
  select(asv_id, otu)

# get names for each otu
rel_abund %>% 
  left_join(otus, by="asv_id") %>% 
  left_join(taxonomy, by="asv_id") %>% 
  distinct(otu, plot_taxa) %>% 
  arrange(otu)

# get rel abund of each otu
rel_abund %>% 
  left_join(otus, by="asv_id") %>% 
  group_by(otu) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  arrange(-rel_abund) %>% 
  select(-rel_abund) %>% 
  add_column(letter = LETTERS[1:9])

# name each OTU, lettered in order of rel abund
otu_names <- tibble(otu = as.character(1:9), 
                    otu_name = c("Chloromonas spp. C",
                                 "Sanguina B",
                                 "Chlainomonas A",
                                 "Chloromonas muramotoi E",
                                 "Chloromonas F",
                                 "Chloromonas H",
                                 "Algae G",
                                 "Trebouxiophyceae I",
                                 "Chloromonas krienitzii D"))

otus_asv_name <- otus %>% 
  left_join(otu_names, by="otu") %>% 
  select(asv_id, otu_name)
```

```{r echo=F, fig.width=8, fig.height=6}
tsne_edited_otus %>% 
  select(plot_taxa, otu) %>% 
  distinct() %>% 
  arrange(otu)

@@

ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), alpha=0.3, show.legend = FALSE) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_otu_names, is.na(plot_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_otu_names, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few() 
```

```{r include_rel_abund, echo=F}
total_rel_abund <- rel_abund %>% 
  group_by(asv_id) %>% 
  summarise(global_rel_abund = sum(rel_abund))

@@

final_tsne <- ggplot(tsne_rel_abund, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), 
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank", 
                             expand = unit(2, "mm")) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
             aes(size = global_rel_abund),
             color = "grey",alpha = 0.5) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order),
                 size = global_rel_abund), 
             alpha = 0.5) +
  labs(color="ASV taxonomic assignment") +
  ggthemes::theme_few() +
  scale_size(range = c(1, 6), guide="none") + ################################################# not changing the scale size
  theme(legend.position = "none")
final_tsne
```

```{r asv_mds, echo=F}
mds_coords <- dists %>% 
  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
  as_tibble() %>% 
  add_column(asv_id = colnames(dists), .before=1)

@@

mds <- ggplot(mds_asv_plot, aes(V1, V2)) + 
  ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
                               filter(otu_name != "Trebouxiophyceae J"), # dont plot the ellipse around other treb
                             aes(fill = otu_name, label=otu_name),
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank",
                             expand = unit(2, "mm")) +
  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
  scale_size(range = c(1, 6), guide="none") +
  scale_alpha(guide="none")+
  theme_few() +
  labs(color="ASV taxonomic assignment")
  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
mds
```

@@

Run IQtree
```{r tree, fig.width=8, fig.height=20}
# root tree with Trebouxiophyceae
root <- taxonomy %>% 
  filter(class=="Trebouxiophyceae") %>% 
  head(1) %>% 
  pull(asv_id)

rbcl.tree <- rbcl.tree %>% root(outgroup = root, resolve.root=T)

@@

# Plot trees

# tips colored by taxonomy
taxonomy_tree <- rbcl_ggtree %<+% taxonomy + 
  geom_tippoint(aes(color=plot_taxa, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

# clades annotated by OTU
otu_tree <- rbcl_ggtree %<+% asv_otu_key + 
  geom_tippoint(aes(color=otu_name, alpha=0.7), size=3) + # 
  geom_text2(data=rbcl_ggtree_data, aes(label=paste0(boot,";",bayes)),  vjust= 0.01, hjust = 1, size=3) + # add bootstrap vals
  theme(legend.position = c(0.2,0.7), 
        legend.title = element_blank(), 
        legend.key = element_blank())

otu_tree

todo: 
- better color choices
- circular tree

```{r output}
write_csv(otus_asv_name, path = here("output/asv_otu_key.csv"))

ggsave(filename = here("manuscript/figs/2_taxonomy/tsne.png"), plot=final_tsne, height = 6, width = 8)
ggsave(filename = here("manuscript/figs/2_taxonomy/mds.png"), plot=mds,  height = 6, width = 8)
ggsave(filename = here("manuscript/figs/tree.png"), plot=otu_tree, width=8, height = 20)
```
